---
topic: "plants-and-animals"
title: "Status of Invasive Species in B.C."
output: envreportbc::print_ver
---

```{r set-options, echo=FALSE, cache = FALSE, warning=FALSE, message=FALSE}
#load libraries for script
library(dplyr) #data munging
library(magrittr) #pipe
library(ggplot2) #plotting
library(scales) # plotting
library(envreportutils) #multiplot function
library(rphylopic) #for black rabbit image
library(knitr) #for knitting to PDF
library(extrafont) #for embed_fonts function
library(rgeos) #for mapping
library(maptools)

opts_chunk$set(echo = FALSE, cache = FALSE, warning = FALSE, message = FALSE, 
               error = FALSE)

opts_knit$set(root.dir = "c:/_dev/invasives/")

## Species summary
read_chunk("../species_counts.R")

## Plants
load("../tmp/plant_analyses.rda")
read_chunk("../plants/03_output.R")

## Aquatics
load("../tmp/aquatics_analyses.rda")
read_chunk("../aquatics/04_output.R")

font_reset <- theme(text = element_text(family = NULL))

```



Invasive species are alien or exotic plants, animals and other organisms
that have been introduced into a natural environment where they are not
normally found and where they negatively impact the environment, people,
or the economy. For example, invasive species can threaten native species
and damage habitats, reduce crop yields and forage for grazing, damage
recreational and commmercial fisheries, and pose helath risks to people.
This indicator reports on the status of invasive species in British
Columbia.

-   Almost a thousand alien plant and animal species are known to have
    established in natural environments in B.C. Plants are
    overwhelmingly the most common group of established alien species in
    the province.
-   Not all alien species cause harm and are considered invasive,
    however the degree of potential impact is often not fully
    understood.
-   Aquatic invasive species have been detected in almost all regions of
    the province. Invasive freshwater fish species in particular have
    expanded in both number and geographic area since the early 1900s.
-   Of the over 800 alien plant species established in the province,
    over 175 are known to be invasive and occur at more than 117,000
    locations across the province.
-   Most invasive species are unintentionally introduced by human
    activities, consequently there are more known invasive species in
    areas of the province where there are more people.

One of B.C.'s proactive approaches to managing invasive species is the
[BC Early Detection Rapid Response
(EDRR)](https://www.for.gov.bc.ca/hra/invasive-species/edrr.htm) program
which focusses on detecting and preventing establishment of new invasive
species. There are a few well known stories from the past of intentional
introductions of invasive species such as [scotch
broom](http://ojs.library.ubc.ca/index.php/bcstudies/article/viewFile/345/1875),
[smallmouth bass, and yellow
perch](http://ibis.geog.ubc.ca/biodiversity/efauna/AlienSpeciesinBritishColumbiaHistoricalRecords.html#FishFrogs),
but with public education and programs like
[EDRR](https://www.for.gov.bc.ca/hra/invasive-species/edrr.htm), these
will hopefully be a thing of the past.

### Number of Alien Species in B.C.

-   As of 2014, 978 alien species are known to have established in B.C. Most of 
    those are plants (86.8%) and 8.5% are invertebrates. The number of alien 
    insects reported are limited to those from groups that are presently listed 
    by the [B.C. Conservation Data Centre](http://www.env.gov.bc.ca/cdc/) 
    (dragonflies, tiger beetles, butterflies and grasshoppers).
-   Birds, reptiles, amphibians, fish and mammals make up 4.6% of the
    known established alien species, including the familiar eastern grey
    squirrel and American bullfrog. The remainder is one fungus species
    (0.1%).
    
```{r species-counts-pre}
```

```{r alien.totals, fig.height=5.5, fig.width=12}
```

\\
\begin{center}
\includegraphics{images/squirrel_ds.jpg}\\*
Eastern grey squirrel (\textit{Sciurus carolinensis})\\*
Photo credit: Dominique Sigg
\end{center}
\newpage

### Number and Distribution of Aquatic Invasive Species in B.C.

-   Aquatic invasive species have been detected in almost all ecological
    drainage units (EDUs) in the province. Aquatic invasive plants are
    the most widespread and numerous group, with 101 species occurring
    in 27 of 36 EDUs.
-   There are only two aquatic invasive reptile species in the province,
    and both are turtles: the red-eared (pond) slider, a popular pet
    turtle that is often released into the wild; and the eastern
    snapping turtle.

```{r aquatics-graphs}

```


```{r fig.width=8, fig.height=8}
library(ggplot2)
chart_facet_map + theme(legend.position = "bottom", 
                        legend.key.height = unit(0.8, "cm"), 
                        legend.key.width = unit(2.5, "cm")) +
  font_reset
```

\begin{center}
\includegraphics{images/bullfrog_lr.jpg}\\*
American bullfrog (\textit{Lithobates catesbeianus})\\*
Photo credit: Leah Ramsay
\end{center}

-   Invasive freshwater fish species have expanded in number and
    geographic range since the early 1900s. They have been found in 25
    of 36 EDUs, with 30 different invasive freshwater fish species
    detected in the province.
-   As with many long-term observational data, some of the observed
    increases in invasive freshwater fish may be due to increased survey
    effort over time.

```{r fig.width=8, fig.height=8}
chart_year_map + theme(legend.position = "bottom", 
                       legend.key.height = unit(0.8, "cm"), 
                       legend.key.width = unit(2.5, "cm")) +
  font_reset
```


```{r}
chart_prov_fish_trend + font_reset
```


### Number and Distribution of Invasive Plant Species in B.C.

-   As of 2014, there are over 117,000 locations with recorded invasive
    plant species in B.C.
-   Habitats, or [biogeoclimatic
    zones](https://www.for.gov.bc.ca/hre/becweb/biogeoclimatic%20zones),
    with the most invasive plant sites are the Interior Douglas-fir,
    Coastal Western Hemlock, Interior Cedar - Hemlock, and the
    Sub-Boreal Spruce zones, all of which have more than 10,000 known
    invasive plant locations. High elevation areas have the fewest
    invasive plant location records (12 or less each).
-   There are a total of 175 recorded invasive plants species in B.C.
    The most common invasive plant species are spotted knapweed, Canada
    thistle, common tansy and oxeye daisy, with more than 10,000 known
    locations for each. However, many of the 175 species are much less
    abundant &mdash; for example 36 species are known to occur at 10 or fewer
    locations in the province at this time.

```{r plants-graphs}
```

```{r fig.height = 4}
sitesmap + ggtitle("Number of Known Invasive Plant Locations") + 
  theme(legend.title = element_blank(), 
        legend.position = "right", 
        legend.key.height = unit(4, "cm"),
        plot.margin = unit(c(0,0,0,0),"mm")) +
  font_reset
  
sppmap + ggtitle("Number of Established Invasive Plant Species") + 
  theme(legend.title = element_blank(), 
        legend.position = "right", 
        legend.key.height = unit(4, "cm"),
        plot.margin = unit(c(0,0,0,0),"mm")) +
  font_reset
```

\begin{center}
\includegraphics[width=76mm]{images/Giant_hogweed_FLNRO_sm.jpg}\\*
Giant hogweed (\textit{Heracleum mantegazzianum})
\end{center}

### About the Data Sources

The data summarized for this indicator &mdash; available in the Data section
below &mdash; are from three different B.C. government programs:

-   *Number of Alien Species* data are from the [BC Conservation Data
    Centre (CDC)](http://www.env.gov.bc.ca/atrisk/toolintro.html).
    The total number of alien species for each species group is the sum
    of all known and *established* exotic subtaxa originating from
    outside of B.C. as of 2014. Subtaxa include species, subspecies, populations
    and varieties. The number of alien species reported by the CDC only includes
    information from the better known groups of plants and animals and is
    therefore a minimum estimate. This is particularly true for insects, as only
    species groups for which the CDC has comprehensive lists (dragonflies, tiger
    beetles, butterflies and grasshoppers) are tracked and reported. As
    invertebrate groups are added to the CDC lists, these will be incorporated
    in future updates to this indicator.
-   *Number and Distribution of Aquatic Invasive Species* data are from
    the [Aquatic Invasive Species
    Program](https://www.for.gov.bc.ca/hra/invasive-species/mussels.htm).
    The data are a subset of the CDC exotic species list as it only
    includes aquatic and semi-aquatic taxonomic groups and alien or
    exotic aquatic species that are considered invasive. However, this
    dataset also includes all aquatic invasive species records &mdash;
    established species, species with single or only a few occurance
    records, and species that are native to one part of B.C. but
    invasive in another part of the province.
-   *Number and Distribution of Invasive Plant Species* data are from
    the [Invasive Alien Plant Program
    Application](https://www.for.gov.bc.ca/hra/Plants/application.htm)
    administered by the [B.C. Invasive Plant
    Program](https://www.for.gov.bc.ca/hra/Plants/index.htm). The data
    are a subset of the CDC exotic plant species list as it only
    includes alien plants that are considered invasive and that have
    occurrence records in B.C. Information on management activities for
    many invasive plants are also contained in this database.

### References and Other Useful Links

[**Least Wanted!** Early Detection-Rapid Response Program for
B.C.](https://www.for.gov.bc.ca/hra/invasive-species/edrr.htm)

Click [HERE](http://www.reportinvasives.ca/) to report an invasive
species in B.C.

For more information on invasive species you can visit the [B.C.
government Inter-Ministry Invasive Species Working
Group](http://www.for.gov.bc.ca/hra/invasive-species/index.htm)\
 and the [B.C. Invasive Plant
Program](http://www.for.gov.bc.ca/hra/Plants/index.htm)

[Invasive Species Council of British Columbia
(ISCBC)](http://www.bcinvasives.ca/)

### Data

\*By accessing these datasets, you agree to the license associated with
each file, as indicated in parentheses below.

-   [BC Species and Ecosystems Conservation Status
    Information](http://catalogue.data.gov.bc.ca/dataset/bc-species-and-ecosystems-conservation-status-information)
    (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))
-   [Aquatic Invasive Species of British
    Columbia](http://catalogue.data.gov.bc.ca/dataset/aquatic-invasive-species-of-british-columbia)
    (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))
-   [EAUBC Ecological Drainage
    Units](http://catalogue.data.gov.bc.ca/dataset/eaubc-ecological-drainage-units)
    (License: [B.C. Crown
    Copyright](http://www2.gov.bc.ca/gov/admin/copyright.page))
-   [B.C. Invasive Alien Plant
    Site](http://catalogue.data.gov.bc.ca/dataset/invasive-alien-plant-site)
    (License: [B.C. Crown
    Copyright](http://www2.gov.bc.ca/gov/admin/copyright.page))
-   [Biogeoclimatic Ecosystem Classification (BEC)
    Map](http://catalogue.data.gov.bc.ca/dataset/biogeoclimatic-ecosystem-classification-bec-map)
    (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))

Published and available online at Envronmental Reporting BC (September 2015): http://www.env.gov.bc.ca/soe/indicators/plants-and-animals/invasive_species.html

Email correspondence to: envreportbc@gov.bc.ca
